Epistemic Semantics in Guarded String Models

### 2020-09-27 tex/sdraft.tex
fix typos and bib

  \(
  \begin{array}{lll}
    \sem{\rho\mathbin{:}\sigma}^\varphi & = &
        {\mathcal A}^{\rho\varphi}_{\mathsf{B}}\times{\mathcal A}^{\sigma\varphi}_{\mathsf{B}}\\[.1em]
        \sem{\zeta +\eta}^\varphi & = & \sem{\zeta}^\varphi\cup\sem{\eta}^\varphi\\
        \sem{\zeta \mathbin{\&}  \eta}^\varphi & = & \sem{\zeta}^\varphi\cap\sem{\eta}^\varphi\\
        \sem{\bar{\zeta}}^\varphi & = & {\mathcal A}^{\varphi}_{\mathsf{B}} \times {\mathcal A}^{\varphi}_{\mathsf{B}} \setminus \sem{\zeta}^\varphi\\
  \end{array}  
  \)
  \end{minipage}}
\caption{Syntax of state formulas and syntax and semantics of effect formulas. Effect formulas
  denote relations between atoms. In a state formula, juxtaposition $\rho\,\sigma$ is conjunction.}
\label{effectfigure}
\end{figure}

@@

\begin{example}   \label{interpretation1}
\hspace{-1.5em}\begin{tabular}[t]{ll}
${\bullet}^*t$ & It's tails. \\
$\bullet^*h$ & It's heads. \\  
  $ \bullet^* h \wedge \Box_a \bullet^* h$  & Amy knows it's heads. \\
\end{tabular}
\vspace{1mm}
\hspace{-1.5em}\begin{tabular}[t]{l}
  $\Box_b(\Box_a\bullet^*t + \Box_a\neg\bullet^*t)$\\
 Bob believes Amy knows whether it's tails.  \\
\end{tabular}
\end{example}
xs
{\em Standard Epistemic Modalities}
Using our existential modal primitive $\Diamond_a$, and the dual
encodings of $\Box_a$ and $\wedge$, we can encode the standard modal
operators expressing knowledge $(\K_a)$ and belief $(\B_a)$
as in \ex{encodings}.\footnote{Deeper analyisis of the lexical semantics
  of {\em know} requires adding modeling of presupposition
  \citep{collard2018finite}. The grammar fragment in Section 6 does not model
  the presupposition of {\em know}, except as
  an entailment.}

@@

{\em Axiomatic Classification}. To situate our logic as a modal logic, consieer the soundness of
the standard modal axioms given our semantics \citep{hughes1996new}. Some of
these standard axioms (see \ex{valid-axioms}) hold all the time, i.e. they
are \emph{valid}, and the remaining axioms are valid when $\hat{R}_a$
has a certain shape. The axioms in \ex{conditionally-valid-axioms}
have a nontrivial condition validity condition.

@@

Defining KAT product using Fst's set-lifted string concatenation (denoted by
juxatposition $\texttt{X} \texttt{Y}$) requires more care. Naively concatenating
strings with atoms (Boolean vectors) at both ends doubles atoms at the juncture, and
does not enforce the requisite atom equality.  To implement KAT product, we define
the binary operation \texttt{Cn}, which concatenates strings in the string algebra,
removes strings with non-matching atoms, and then deletes the second of two atoms to
create a set of well-formed guarded strings.  See Figure \ref{fstfigure1}. {\tt Wf0}
is the set of ordinary strings that does not contain unequal pairs of atoms, as
defined using Fst's containment operator \texttt{\$}. The \texttt{Squash} relation
uses Fst's rewrite notation to delete atoms (elements of \texttt{St}) that are
preceded by another atom. \footnote{ This is a non-equal length regular relation. The
  finite state calculus includes such relations, and they can be used with relation
  composition and relation domain and co-domain.  They are restricted in that the
  complement and set difference for non-equal length relations is not defined.
  Epistemic alterative relations are equal-length relations.} This relation is
applied via the relational composition (\texttt{.o.}) and codomain (\texttt{.l})
operators.
%% Deletion is accomplished by a
%% re-write rule in the finite state calculus, which is a notation for
%% defining regular relations by contextually constrained substitutions.
%% In this case, is {\tt Squash} is a regular relation that deletes an
%% atom (a sequence of 0's and 1's of a certain length) when it follows
%% an atom.

@@

\begin{figure} \label{relclosure}
  \begin{tcolorbox}[size=fbox]
  \footnotesize  
{\raggedright
  \begin{hangingpar}
%% {\tt{define RelKpl(R)}}\\
%% $\underbrace{\mbox{\tt{Squash.i\hspace{1pt}.o.}}}_c$
%% $\underbrace{\mbox{\tt Wf0\hspace{1pt}.o.}}_b$
%% $\underbrace{\mbox{\tt{[R+]}}}_a$
%% $\underbrace{\mbox{\tt .o.\hspace{1pt}Wf0}}_b$
%% $\underbrace{\mbox{\tt{.o.\hspace{1pt}Squash}}}_c$
{\tt{define RelKpl(R)}}\\
$\!\underbrace{\texttt{Squash.i}\,\texttt{.o.}}_c
\underbrace{\mbox{\tt Wf0\,.o.}}_b
\underbrace{\mbox{\tt{[R+]}}}_a
\underbrace{\mbox{\tt .o.\,Wf0}}_b
\underbrace{\mbox{\tt{.o.\,Squash}}}_c$

@@

It remains to define an epistemic alternative relation on worlds for each agent.  The
relevant information in Figure \ref{epikfigure} is a relation between bare events for
each agent. This determines a relation between bare events decorated with compatible
atoms. 
In Fst, we use the closure of the concatenation
product operation on relations to lift a relation on decorated events for an agent to
the corresponding relation on worlds.  The concatenation product $RS$ of two
relations $R$ and $S$ is the set of pairs of the form $\pair{x_1x_2}{y_1y_2}$, where
$x_1 \mathrel R y_1$, and $x_2 \mathrel S y_2$.  In Fst, {\tt{R+}} is the closure of relation {\tt R} with
respect to this operation. Figure \ref{relclosure} defines the corresponding
operation on sets of guarded strings as encoded in Fst.\footnote{Relation concatenation in Fst
  differs from relation composition ({\tt{.o.}}), and the closure under discussion here is
  the closure of the former rather than the latter.}  The epistemic alternative
relation on worlds for an agent is then defined as the KAT relational concatenation
closure {\tt RelKst} of the decorated-event alternative relation for the agent.

@@

The bounded interpretation into lists of strings is very similar to
the unboundeds interpretation into sets of strings, except for the
bounds checking. The full details are shown in
Figure~\ref{fig:list-interp}. First note that when $n=0$, the
denotation is empty, denoted $[]$. Terms of the form $0$, $1$, $e$,
and $\psi$ have the same denotation as before, translated into a list
(denoted $\lfloor S\rfloor$, for a set $S$). We compute atoms using
BBDs, which concisely represent boolean functions~\cite{BBDS}.

@@

\end{examples}
Sentence \exi{three}{d} is assigned the logical form \exi{three}{e} by the grammar.
Logical relations between propositions are checked in the finite state calculus by
checking set-theoretic relations between sets of worlds.
For instance entailment $p \rightarrow q$ is decided by checking in an interpreter
for the finite sate calculus whether $p - q$ is non-empty.
In the model defined by Figure 2, the
propositions \exi{three}{b} and \exi{three}{e} are
independent. They are equivalent in a version 
without secret flipping.

@@

Open source code will be distributed in conjunction with the conference.

>>>>

  \(
  \begin{array}{lll}
    \sem{\rho\mathbin{:}\sigma}^\varphi & \triangleq &
        {\mathcal A}^{\rho\varphi}_{\mathsf{B}}\times{\mathcal A}^{\sigma\varphi}_{\mathsf{B}}\\[.1em]
        \sem{\zeta +\eta}^\varphi & \triangleq & \sem{\zeta}^\varphi\cup\sem{\eta}^\varphi\\
        \sem{\zeta \mathbin{\&}  \eta}^\varphi & \triangleq & \sem{\zeta}^\varphi\cap\sem{\eta}^\varphi\\
        \sem{\bar{\zeta}}^\varphi & \triangleq & {\mathcal A}^{\varphi}_{\mathsf{B}} \times {\mathcal A}^{\varphi}_{\mathsf{B}} \setminus \sem{\zeta}^\varphi\\
  \end{array}  
  \)
  \end{minipage}}
\caption{Syntax of state formulas and syntax and semantics of effect formulas. Effect formulas
  denote relations between atoms. In a state formula, juxtaposition $\rho\,\sigma$ is conjunction.}
\label{effectfigure}
\end{figure}

@@

\begin{example}   \label{interpretation1}
\hspace{-1.5em}\begin{tabular}[t]{ll}
${\bullet}^*t$ & It's tails. \\
$\bullet^*h$ & It's heads. \\  
  $ \bullet^* h \wedge \Box_a \bullet^* h$  & Amy knows it's heads. \\
\end{tabular}
\vspace{1mm}
\hspace{-1.5em}\begin{tabular}[t]{l}
  $\Box_b(\Box_a\bullet^*t + \Box_a\neg\bullet^*t)$\\
 Bob believes Amy knows whether it's tails.  \\
\end{tabular}
\end{example}

{\em Standard Epistemic Modalities}
Using our existential modal primitive $\Diamond_a$, and the dual
encodings of $\Box_a$ and $\wedge$, we can encode the standard modal
operators expressing knowledge $(\K_a)$ and belief $(\B_a)$
as in \ex{encodings}.\footnote{Deeper analyisis of the lexical semantics
  of {\em know} requires adding modeling of presupposition
  \citep{collard2018finite}. The grammar fragment in Section 6 does not model
  the presupposition of {\em know}, except as
  an entailment.}

@@

{\em Axiomatic Classification}. To situate our logic as a modal logic, consider the soundness of
the standard modal axioms given our semantics \citep{hughes1996new}. Some of
these standard axioms (see \ex{valid-axioms}) hold all the time, i.e. they
are \emph{valid}, and the remaining axioms are valid when $\hat{R}_a$
has a certain shape. The axioms in \ex{conditionally-valid-axioms}
have a nontrivial condition validity condition.

@@

Defining KAT product using Fst's set-lifted string concatenation (denoted by
juxtaposition $\texttt{X}~\texttt{Y}$) requires more care. Naively concatenating
strings with atoms (Boolean vectors) at both ends doubles atoms at the juncture, and
does not enforce the requisite atom equality.  To implement KAT product, we define
the binary operation \texttt{Cn}, which concatenates strings in the string algebra,
removes strings with non-matching atoms, and then deletes the second of two atoms to
create a set of well-formed guarded strings.  See Figure \ref{fstfigure1}. {\tt Wf0}
is the set of ordinary strings that does not contain unequal pairs of atoms, as
defined using Fst's containment operator \texttt{\$}. The \texttt{Squash} relation
uses Fst's rewrite notation to delete atoms (elements of \texttt{St}) that are
preceded by another atom. \footnote{ This is a non-equal length regular relation. The
  finite state calculus includes such relations, and they can be used with relation
  composition and relation domain and co-domain.  They are restricted in that the
  complement and set difference for non-equal length relations is not defined.
  Epistemic alterative relations are equal-length relations.} This relation is
applied via the relational composition (\texttt{.o.}) and codomain (\texttt{.l})
operators.
%% Deletion is accomplished by a
%% re-write rule in the finite state calculus, which is a notation for
%% defining regular relations by contextually constrained substitutions.
%% In this case, is {\tt Squash} is a regular relation that deletes an
%% atom (a sequence of 0's and 1's of a certain length) when it follows
%% an atom.

@@

\begin{figure}
  \begin{tcolorbox}[size=fbox]
  \footnotesize  
{\raggedright
  \begin{hangingpar}
%% {\tt{define RelKpl(R)}}\\
%% $\underbrace{\mbox{\tt{Squash.i\hspace{1pt}.o.}}}_c$
%% $\underbrace{\mbox{\tt Wf0\hspace{1pt}.o.}}_b$
%% $\underbrace{\mbox{\tt{[R+]}}}_a$
%% $\underbrace{\mbox{\tt .o.\hspace{1pt}Wf0}}_b$
%% $\underbrace{\mbox{\tt{.o.\hspace{1pt}Squash}}}_c$
{\tt{define RelKpl(R)}}\\
$\!\underbrace{\texttt{Squash.i}\,\texttt{.o.}}_c
\underbrace{\mbox{\tt Wf0\,.o.}}_b
\underbrace{\mbox{\tt{[R+]}}}_a
\underbrace{\mbox{\tt .o.\,Wf0}}_b
\underbrace{\mbox{\tt{.o.\,Squash}}}_c$

@@

It remains to define an epistemic alternative relation on worlds for each agent.  The
relevant information in Figure \ref{epikfigure} is a relation between bare events for
each agent. This determines a relation between bare events decorated with compatible
atoms. 
In Fst, we use the closure of the concatenation
product operation on relations to lift a relation on decorated events for an agent to
the corresponding relation on worlds.  The concatenation product $R~S$ of two
relations $R$ and $S$ is the set of pairs of the form $\pair{x_1x_2}{y_1y_2}$, where
$x_1 \mathrel R y_1$, and $x_2 \mathrel S y_2$.  In Fst, {\tt{R+}} is the closure of relation {\tt R} with
respect to this operation. Figure \ref{relclosure} defines the corresponding
operation on sets of guarded strings as encoded in Fst.\footnote{Relation concatenation in Fst
  differs from relation composition ({\tt{.o.}}), and the closure under discussion here is
  the closure of the former rather than the latter.}  The epistemic alternative
relation on worlds for an agent is then defined as the KAT relational concatenation
closure {\tt RelKst} of the decorated-event alternative relation for the agent.

@@

The bounded interpretation into lists of strings is very similar to
the unboundeds interpretation into sets of strings, except for the
bounds checking. The full details are shown in
Figure~\ref{fig:bound-interp}. First note that when $n=0$, the
denotation is empty, denoted $[]$. Terms of the form $0$, $1$, $e$,
and $\psi$ have the same denotation as before, translated into a list
(denoted $\lfloor S\rfloor$, for a set $S$). We compute atoms using
BDDs, which concisely represent boolean functions~\cite{lee1959representation}.

@@

\end{examples}
Sentence \exi{three}{d} is assigned the logical form \exi{three}{e} by the grammar.
Logical relations between propositions are checked in the finite state calculus by
checking set-theoretic relations between sets of worlds.
For instance entailment $p \rightarrow q$ is decided by checking in an interpreter
for the finite state calculus whether $p - q$ is non-empty.
In the model defined by Figure 2, the
propositions \exi{three}{b} and \exi{three}{e} are
independent. They are equivalent in a version 
without secret flipping.

@@

We will release Epik's source code under an open source license prior
to the conference.
